{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-08-06 submit/uist25-52/Source/paper.tex
Fix keyidea boxes and circled letters
Tweak table widths
fix orphan in intro
Adjust margin fig 3
Page break tweaks, margins, circled hack

\newcommand*\circled[1]
{{\Large \textcircled{\footnotesize {#1}}}}

% \DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
%   boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
%   colback=gray!20,colframe=gray!20,
%   enlarge bottom by=0pt,enlarge top by=0pt,
%   arc=0pt,outer arc=0pt]
% \lettrine[lraise=0.3]{\LARGE #1}{~}
% \small \textbf{#2.} #3
% \end{tcolorbox}
% }
\DeclareRobustCommand{\keyideabox}[3]
{\vspace{\dimexpr\baselineskip\relax} \colorbox{gray!20}{
\parbox{\dimexpr\columnwidth-\marginparsep+1pt\relax}
{\small {#1} \textbf{#2.} #3}
}}

@@

\begin{itemize}
\item \emph{Collaborative Editing.} Users should be able to locally modify a shared document
  and merge their changes, preferably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations for working with edit histories.
\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
  implemented on top of the Denicek substrate (\S\ref{sec:impl}).
\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
  The analysis shows that the desired functionality requires a careful choice among interconnected
  design options.
\item To evaluate how Denicek simplifies the development of program\-ming systems, we
  build an innovative data science notebook Datnicek and assess its implementation
  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system (\S\ref{sec:eval}).
\end{itemize}

@@

\subsubsection*{\circled{A} Adding a Speaker}
Evelyn starts with an empty document, which is represented as a record. She adds a field for
each heading and a field named {\small\texttt{speakers}}
containing a list {\small\texttt{<ul>}}. \diff{She
uses the command toolbox to issue edit commands that add the first speaker.}

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
the details of another speaker \diff{into the textbox}, adds a new {\small\texttt{<li>}} element
and copies the speaker details from the textbox into the new element in the source view,
\diff{using a copy command}.

\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker,
Evelyn opens the history view, selects edits that added the speaker and saves them as the
{\small\texttt{add-speaker}} interaction. She attaches this interaction as an event handler for the
{\small\texttt{click}} event of the button.

@@

\subsubsection*{\circled{D} Refactoring Document Structure} Alfred starts with the initial
version of the document (A) and turns the list into a table. He invokes a series of commands that
\diff{rename tags, wrap elements, copy values and split strings using the comma as a separator}.
\note{More concrete description of edits}

\subsubsection*{\circled{E} Merging Edits}
\diff{Alfred then merges the later edits (B, C) done by Evelyn into his version of the document.}
The refactoring is applied to all speakers. New speakers added using the ``Add speaker!''
button are also automatically transformed to the new format.

@@

\subsubsection*{\circled{F} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
the initial document. \diff{She uses the command toolbox to create document nodes representing
formulas. Formulas are regular nodes with a special} {\small\texttt{x-formula}} \diff{tag and
arguments as child nodes. To count the speakers, she uses the} {\small\texttt{count}} \diff{builtin
with a reference to the node representing the speaker list as the argument.}\note{Modified to describe
representation of formulas.}

@@

\subsubsection*{\circled{G} Merging Formulas} When Juliana merges the budget calculation (F) with
the earlier edits (E), references in formulas are automatically updated to point to the \diff{rows of the table}.
Adding new speaker via the ``Add speaker!'' button \diff{invalidates results of only those formulas
that depend on the number of speakers.}

@@

\begin{tabular}{|llp{18.08em}|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
\sffamily\small{\bfseries Selector} & {\sffamily\footnotesize Notation} & \\[0.2em]
\hline
&&\\[-1em]
\seltablecol{\texttt{..}}{Parent}{Refers to a parent of a node}
\seltablecol{\texttt{field}}{Field}{Refers to record field of a given name}
\seltablecol{texttt{\textbackslash\#index}}{Index}{Refers to list element at a given index}
\seltablecol{\texttt{*}}{Any}{Refers to all children of a list node}
\hline
\end{tabular}

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \texttt{<tag>} with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \texttt{<tag>} with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \texttt{selectors} as a link.}
\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\end{figure}

@@

\item \emph{Transform Matching References.} If $e_2$ targets a node that is inside a node
  whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
  corresponds to the new structure. This is done using the rules shown in Fig.~\ref{fig:updates},
  that apply when transforming references inside a document, although we now also support the case
  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs} for details).
\end{enumerate}

@@

\begin{figure}
\newcommand{\tttablecol}[5]{
\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
}
\begin{tabular}{|p{27em}|}
\hline
\\[-1em]
\tttablecol{RenameField}{target, old field, new field}{\texttt{/target/old\_field/nested}}{\texttt{/target/new\_field/nested}}
  {Replace Field for matching references.}
\tttablecol{WrapRecord}{target, tag, field}{\texttt{/target/nested}}{\texttt{/target/field/nested}}
  {Insert extra Field selector after matching prefix.}
\tttablecol{WrapList}{target, index, tag}{\texttt{/target/nested}}{\texttt{/target/*/nested}}
  {Insert extra All selector after matching prefix.}
\hline
\end{tabular}
\vspace{-0.75em}
\caption{How document edits transform references}
\label{fig:updates}
\vspace{-0.5em}
\end{figure}

@@

\begin{figure}
\newcommand{\eftablecol}[3]{
\small{\bfseries #1}\;\footnotesize\textit{target}\,\; --\,\; {\footnotesize #2}\\[-0.1em]
\quad \small #3\\[0.3em]
}
\begin{tabular}{|p{27em}|}
\hline
\\[-1em]
\eftablecol{StructureEffect}{Affects fields or structure of the target node.}
  {\ident{RenameField}, \ident{DeleteField}, \ident{WrapRecord}, \ident{WrapList}, \ident{Copy}}
\eftablecol{ValueEffect}{Transforms value, modifies list or adds an additional field.}
  {\ident{Add}, \ident{Append}, \ident{Reorder}, \ident{DeleteItem}, \ident{PrimitiveEdit}}
\eftablecol{TagEffect}{Modifies the tag of the target node.}
  {\ident{UpdateTag}}
\hline
\end{tabular}
\vspace{-0.75em}
\caption{Effects of individual edit operations}
\label{fig:effects}
\vspace{-1em}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
\vspace{-1em}
\caption{Using provenance tracking to highlight document parts that contributed to the
calculation of refreshments costs (1), travel costs (2) and all costs (3).}
\label{fig:provenance}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
\label{fig:coevolution}
\end{figure}

@@

\begin{itemize}
\item \emph{Structure Editing.} Code in code cells should be edited through structure editor
  as in Histogram \cite{petricek-2019-histogram}. This allows Datnicek to keep track of the code
  structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
  We do not initially aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.

@@

\item \emph{Code Completion.} During editing, code completion should offer available data
  transformations and operations as when using type providers \cite{syme-2013-inforich} or
  iterative prompting in The Gamma \cite{petricek-2022-thegamma}.

@@

\item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
  convertible into code that can be checked by the data scientist and further edited as in Wrex \cite{drossos-2020-wrex}.
\end{itemize}

@@

\begin{acks}
The authors thank Clemens Nylandsted Klokmose for research guidance and
attendees of the IFIP WG 2.16 meeting in Serpiano, members of the Software Architecture Group at HPI
and attendees of LIVE 2023 for feedback on the project.
Anonymous reviewers provided invaluable actionable guidance for improving the final version
of the paper. The work has been supported by the
\grantsponsor{cuni}{Charles University}{https://d3s.mff.cuni.cz/projects/primus24/}
grant \grantnum{cuni}{PRIMUS/24/SCI/021} and by the
\grantsponsor{erccz}{Czech Ministry of Education, Youth and Sports}{https://starfos.tacr.cz/en/projekty/LL2325}
grant ERC-CZ \grantnum{erccz}{LL2325}.
\end{acks}

@@

\begin{figure*}
\newcommand{\extablecol}[5]{
\small{\bfseries #1} #2\; --\; #3\\
\quad \footnotesize #4 \\[-0.2em]
\quad \footnotesize Programming experiences: \emph{#5} \\[0.6em]
}
\begin{tabular}{|l|}
\hline
\\[-1em]
\extablecol{Counter App}{\cite{kiss-2014-7guis}}
  {Counter with increment and decrement buttons.}
  {The current count is represented by a formula that is modified by the buttons.
  The user can inspect the evaluation trace to see how the count was modified.}
  {Programming by Demonstration, Incremental Recomputation, End-User Debugging}
\extablecol{Todo App}{\cite{osmani-2024-todomvc}}
  {Todo with buttons to add an item and remove all completed items.}
  {Adding an item must correctly merge with independently added functionality to
   compute which items are completed and remove them based on a formula result.}
  {Collaborative Editing, Programming by Demonstration, Incremental Recomputation, Schema Change Control}
\extablecol{Conference List}{\cite{edwards-2025-schema}}
  {Managing a list of invited conference speakers and schema change.}
  {Adding speakers to a list through an in-document user interface merges with
   refactoring that turns the list into a table and separates name from an email.}
  {Collaborative Editing, Programming by Demonstration, Schema Change Control}
\extablecol{Conference Budget}{\cite{edwards-2025-schema}}
  {Calculate budget based on a speaker list or a table.}
  {References are updated  when the list is refactored. Only affected formulas
   are recomputed and the user can view elements on which the result depends.}
  {Collaborative Editing, Incremental Recomputation, Schema Change Control, End-User Debugging}
\extablecol{Hello World}{\cite{miller-2001-simult}}
  {Normalize the capitalization of two word messages.}
  {An operation to normalize the text in a list item can be recorded and applied to
   all list items or, alternatively, applied directly to all list items.}
  {Programming by Demonstration}
\extablecol{Traffic Accidents}{\cite{edwards-2022-copypaste}}
  {Compute statistics using two data sources.}
  {Formula to compute statistics can be reused with a different data source;
   error correction is propagated automatically to the copied version of the formula.}
  {Concrete Programming, Incremental Recomputation}
\hline
\end{tabular}

\caption{Formative examples used in Denicek design}
\label{fig:examples}
\end{figure*}

@@

>>>>

\newcommand*\circled[1]
{{\normalfont \Large \textcircled{\footnotesize {#1}}}}

% \DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
%   boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
%   colback=gray!20,colframe=gray!20,
%   enlarge bottom by=0pt,enlarge top by=0pt,
%   arc=0pt,outer arc=0pt]
% \lettrine[lraise=0.3]{\LARGE #1}{~}
% \small \textbf{#2.} #3
% \end{tcolorbox}
% }
\DeclareRobustCommand{\keyideabox}[3]
{\vspace{\dimexpr0.5\baselineskip\relax} \noindent\setlength\fboxsep{6pt}\colorbox{gray!20}{%
\parbox{\dimexpr\columnwidth-\marginparsep\relax}
{\small {#1}\; \textbf{#2.} #3}
}}

@@

\begin{itemize}
\item \emph{Collaborative Editing.} Users should be able to locally modify a shared document
  and merge their changes in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
\item \emph{Programming by Demonstration.} Allow users to construct programs by
  enacting the steps of the desired behavior using concrete examples and generalizing from those~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask about provenance~\cite{cheney-2009-provenance} to understand
  why a computation resulted in a particular result and what inputs contributed to it \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations for working with edit histories.
\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
  implemented on top of the Denicek substrate (\S\ref{sec:impl}).
\item We analyse key design decisions, alternatives and limitations (\S\ref{sec:discuss}), showing
  that the desired functionality requires a careful choice among interconnected
  design options.
\item To evaluate how Denicek simplifies the development of program\-ming systems, we
  build an innovative data science notebook and assess its implementation
  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system~(\S\ref{sec:eval}).
\end{itemize}

@@

\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=0em 0em 30mm 0em]{fig/circled.pdf}} Adding a Speaker}
Evelyn starts with an empty document, which is represented as a record. She adds a field for
each heading and a field named {\small\texttt{speakers}}
containing a list {\small\texttt{<ul>}}. \diff{She
uses the command toolbox to issue edit commands that add the first speaker.}

\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=5mm 0em 25mm 0em]{fig/circled.pdf}} Creating a User Interface} To simplify adding of
further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
the details of another speaker \diff{into the textbox}, adds a new {\small\texttt{<li>}} element
and copies the speaker details from the textbox into the new element in the source view,
\diff{using a copy command}.

\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=10mm 0em 20mm 0em]{fig/circled.pdf}} Abstracting the Interaction} After adding the speaker,
Evelyn opens the history view, selects edits that added the speaker and saves them as the
{\small\texttt{add-speaker}} interaction. She attaches this interaction as an event handler for the
{\small\texttt{click}} event of the button.

@@

\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=15mm 0em 15mm 0em]{fig/circled.pdf}} Refactoring Document Structure} Alfred starts with the initial
version of the document (A) and turns the list into a table. He invokes a series of commands that
\diff{rename tags, wrap elements, copy values and split strings using the comma as a separator}.
\note{More concrete description of edits}

\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=20mm 0em 10mm 0em]{fig/circled.pdf}} Merging Edits}
\diff{Alfred then merges the later edits (B, C) done by Evelyn into his version of the document.}
The refactoring is applied to all speakers. New speakers added using the ``Add speaker!''
button are also automatically transformed to the new format.

@@

\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=25mm 0em 5mm 0em]{fig/circled.pdf}} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
the initial document. \diff{She uses the command toolbox to create document nodes representing
formulas. Formulas are regular nodes with a special} {\small\texttt{x-formula}} \diff{tag and
arguments as child nodes. To count the speakers, she uses the} {\small\texttt{count}} \diff{builtin
with a reference to the node representing the speaker list as the argument.}\note{Modified to describe
representation of formulas.}

@@

\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=30mm 0em 0mm 0em]{fig/circled.pdf}} Merging Formulas} When Juliana merges the budget calculation (F) with
the earlier edits (E), references in formulas are automatically updated to point to the \diff{rows of the table}.
Adding new speaker via the ``Add speaker!'' button \diff{invalidates results of only those formulas
that depend on the number of speakers.}

@@

\begin{tabular}{|llp{16em}|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
\sffamily\small{\bfseries Selector} & {\sffamily\footnotesize Notation} & \\[0.2em]
\hline
&&\\[-1em]
\seltablecol{\texttt{..}}{Parent}{Refers to a parent of a node}
\seltablecol{\texttt{field}}{Field}{Refers to record field of a given name}
\seltablecol{texttt{\textbackslash\#index}}{Index}{Refers to list element at a given index}
\seltablecol{\texttt{*}}{Any}{Refers to all children of a list node}
\hline
\end{tabular}

@@

\begin{tabular}{|cp{23.48em}|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \texttt{<tag>} with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \texttt{<tag>} with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \texttt{selectors} as a link.}
\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\end{figure}

@@

\item \emph{Transform Matching References.} If $e_2$ targets a node that is inside a node
  whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
  corresponds to the new structure. This is done using the rules in Fig.~\ref{fig:updates},
  that apply when transforming references inside a document, although we now also support the case
  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs}).
\end{enumerate}

@@

\begin{figure}
\newcommand{\tttablecol}[5]{
\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
}
\begin{tabular}{|p{25.5em}|}
\hline
\\[-1em]
\tttablecol{RenameField}{target, old, new}{\texttt{/target/old\_field/nested}}{\texttt{/target/new\_field/nested}}
  {Replace Field for matching references.}
\tttablecol{WrapRecord}{target, tag, field}{\texttt{/target/nested}}{\texttt{/target/field/nested}}
  {Insert Field selector after matching prefix.}
\tttablecol{WrapList}{target, index, tag}{\texttt{/target/nested}}{\texttt{/target/*/nested}}
  {Insert extra All selector after matching prefix.}
\hline
\end{tabular}
\vspace{-0.75em}
\caption{How document edits transform references}
\label{fig:updates}
\vspace{-0.5em}
\end{figure}

@@

\begin{figure}
\newcommand{\eftablecol}[3]{
\small{\bfseries #1}\;\footnotesize\textit{target}\,\; --\,\; {\footnotesize #2}\\[-0.1em]
\quad \small #3\\[0.3em]
}
\begin{tabular}{|p{25.5em}|}
\hline
\\[-1em]
\eftablecol{StructureEffect}{Affects fields or structure of the target node.}
  {\ident{RenameField}, \ident{DeleteField}, \ident{WrapRecord}, \ident{WrapList}, \ident{Copy}}
\eftablecol{ValueEffect}{Transforms value, modifies list or adds a field.}
  {\ident{Add}, \ident{Append}, \ident{Reorder}, \ident{DeleteItem}, \ident{PrimitiveEdit}}
\eftablecol{TagEffect}{Modifies the tag of the target node.}
  {\ident{UpdateTag}}
\hline
\end{tabular}
\vspace{-0.75em}
\caption{Effects of individual edit operations}
\label{fig:effects}
\vspace{-1em}
\end{figure}

@@

\begin{figure}[h]
\vspace{0.5em}
\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
\caption{Using provenance tracking to highlight document parts that contributed to the
calculation of refreshments costs (1), travel costs (2) and all costs (3).}
\label{fig:provenance}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
\label{fig:coevolution}
\vspace{0.5em}
\end{figure}

@@

\begin{itemize}
\item \emph{Structure Editing.} Code in code cells should be edited through structure editor
  as in Histogram \cite{petricek-2019-histogram}. This allows Datnicek to keep track of the code
  structure and can make the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
  We do not initially aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.

@@

\item \emph{Code Completion.} During editing, code completion should offer available data
  transformations and operations as when using type providers \cite{syme-2013-inforich} or
  when using iterative prompting in The~Gamma \cite{petricek-2022-thegamma}.

@@

\item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
  convertible into code that can be checked and further edited as in Wrex \cite{drossos-2020-wrex}.
\end{itemize}

@@

\begin{acks}
The authors thank Clemens Nylandsted Klokmose for research guidance,
attendees of the IFIP WG 2.16 meeting in Serpiano, members of the Software Architecture Group at HPI
and attendees of LIVE 2023 for feedback on the project.
Anonymous reviewers provided invaluable actionable guidance for improving the final version
of the paper. The work has been supported by the
\grantsponsor{cuni}{Charles University}{https://d3s.mff.cuni.cz/projects/primus24/}
grant \grantnum{cuni}{PRIMUS/24/SCI/021} and by the
\grantsponsor{erccz}{Czech Ministry of Education, Youth and Sports}{https://starfos.tacr.cz/en/projekty/LL2325}
grant ERC-CZ \grantnum{erccz}{LL2325}.
\end{acks}

@@

@@

\begin{figure}[t]
\newcommand{\extablecolx}[6]{\hline\\[-0.8em]
\small{\bfseries #1} #2\\[-0.1em] \small #3\\[0.3em]
\quad \parbox{7cm}{\footnotesize #4}\\
\quad \parbox{7cm}{\vspace{0.5em}\footnotesize #6: \emph{#5}} \\[1em]
}
\newcommand{\extablecol}[5]{\extablecolx{#1}{#2}{#3}{#4}{#5}{Programming experiences}}
\begin{tabular}{|l|}
\extablecol{Counter App}{\cite{kiss-2014-7guis}}
  {Counter with increment and decrement buttons.}
  {The current count is represented by a formula that is modified by the buttons.
  Formula modification is shown once and then repeated on click.
  The user can inspect the evaluation trace to see how the count was modified.}
  {Programming by Demonstration, Incremental Recomputation, End-User Debugging}
\extablecol{Todo App}{\cite{osmani-2024-todomvc}}
  {Todo with buttons to add an item and remove all completed items.}
  {Functionality to add an item is shown once and then recorded.
   Adding an item must correctly merge with independently added functionality to
   compute which items are completed and remove them based on a formula result.}
  {Collaborative Editing, Programming by Demonstration, Incremental Recomputation, Schema Change Control}
\extablecol{Conference List}{\cite{edwards-2025-schema}}
  {Managing a list of invited conference speakers and schema change.}
  {Adding speakers to a list through an in-document user interface merges with
   refactoring that turns the list into a table and separates name from an email by
   copying the value and removing a part of the string before/after a comma.}
  {Collaborative Editing, Programming by Demonstration, Schema Change Control}
\extablecol{Conference Budget}{\cite{edwards-2025-schema}}
  {Calculate budget based on a speaker list or a table.}
  {References are updated  when the list is refactored. Only affected formulas
   are recomputed and the user can view elements on which the result depends.}
  {Collaborative Editing, Incremental Recomputation, Schema Change Control, End-User Debugging}
\extablecol{Hello World}{\cite{miller-2001-simult}}
  {Normalize the capitalization of two word messages.}
  {An operation to normalize the text in a list item can be recorded and applied to
   all list items or, alternatively, applied directly to all list items.}
  {Programming by Demonstration}
\extablecolx{Traffic Accidents}{\cite{edwards-2022-copypaste}}
  {Compute statistics using two data sources.}
  {Formula to compute statistics can be reused with a different data source. If an error
   in the formula is corrected, the correction is propagated automatically to the copied
   version of the formula.}
  {Concrete Programming, Incremental Recomputation}{Experiences}
\hline
\end{tabular}

\caption{Formative examples used in Denicek design}
\label{fig:examples}
\end{figure}
